Verification of the CVM algorithm with a New Recursive Analysis Technique

### 2025-02-23 paper/sec/intro.tex
minor fixes (mostly macros or uniformity)

We have also mechanized the transformation-based CVM proof by Chakraborty et al.~\cite{chakraborty2022,chakraborty2023}.
As a rough point of comparison, verification of the CVM algorithm using our new technique requires only 1044 lines in Isabelle~\cite{nipkow2002}, while the original proof required \todo{x} lines.

The organization of this paper is as follows.
\cref{sec:background} provides background information on randomized algorithms, in particular on their semantics in Isabelle~\cite{nipkow2002}.
\cref{sec:invariants} introduces our new technique and explains how probabilistic loop invariants can be used to establish tail bounds for the original CVM algorithm.
\cref{sec:negdep} introduces the concept of negative association and our new total and unbiased variant of the CVM algorithm.
\cref{sec:formalization} presents the formalization of both variants of the algorithm, and \cref{sec:formalization_neg_dep} describes our new formalized library on negatively associated random variables.
In \cref{sec:transformation_based_proof}, we discuss some challenges faced in our alternative verification of CVM using the transformation-based proof by Chakraborty et al.~\cite{chakraborty2023}.
The final sections conclude with an overview of related work and a summary of our results.

>>>>

We have also mechanized the transformation-based CVM proof by Chakraborty et al.~\cite{chakraborty2022,chakraborty2023}.
As a rough point of comparison, verification of the CVM algorithm using our new technique requires only 1044 lines in Isabelle/HOL~\cite{nipkow2002}, while the original proof required \todo{x} lines.

The organization of this paper is as follows.
\cref{sec:background} provides background information on randomized algorithms, in particular on their semantics in Isabelle/HOL~\cite{nipkow2002}.
\cref{sec:invariants} introduces our new technique and explains how probabilistic loop invariants can be used to establish tail bounds for the original CVM algorithm.
\cref{sec:negdep} introduces the concept of negative association and our new total and unbiased variant of the CVM algorithm.
\cref{sec:formalization} presents the formalization of both variants of the algorithm, and \cref{sec:formalization_neg_dep} describes our new formalized library on negatively associated random variables.
In \cref{sec:transformation_based_proof}, we discuss some challenges faced in our alternative verification of CVM using the transformation-based proof by Chakraborty et al.~\cite{chakraborty2023}.
The final sections conclude with an overview of related work and a summary of our results.

### 2025-02-23 paper/sec/invariants.tex
minor fixes (mostly macros or uniformity)

An alternative is the Cram\'{e}r--Chernoff method, which is a general method to obtain tail bounds for any random variable.
It can be stated simply as $P(X \geq a) \leq M(t) e^{-ta}$ for all $t > 0$, where $M(t) := \expect [\exp(t X)]$ is the moment generating function of $X$.
%\footnote{The moment generating function is sometimes only defined for some values of $t$ (when the corresponding integral exists), and this is typically an interval including $0$; the inequality is only applicable wherever $M(t)$ is defined.}
%Since the probability space representing the state of the CVM algorithm is finite, $M(t)$ will always be defined in our case.
It is also possible to obtain lower tail bounds $P(X \leq a)$ using Cram\'{e}r--Chernoff method, which just requires estimates for $M(t)$ for $t < 0$, instead of $t > 0$.

@@

Of course, the exponential function in $M(t)$ is convex, so this is not yet the full story.
However, we can instead try to derive tail bounds for the random variable $\indicat(p \geq q) p^{-1} |\chi|$.
This leads to a similar inequality:
\begin{align}
  \expect \left[ \prod_{s \in A} I(p \geq q) h(p^{-1} X_s) \right] \leq h(1)^{|A|} \label[ineq]{i:func_invariant_capped}
\end{align}
with the new condition that $h$ needs to be non-negative and concave only on $[0;q^{-1}]$.
This then allows us to use approximate the exponential function from above with an affine function $h$ on the range $[0;q^{-1}]$, which yields tail bounds for $p^{-1} |\chi|$ under the condition $p \geq q$. As an example, the upper tail bound can be derived as follows:
\begin{eqnarray*}
  \prob( p^{-1} |\chi| \geq (1+\varepsilon) |A| \wedge p \geq q) & \leq &
  \prob( I(p \geq q) p^{-1} |\chi| \geq (1+\varepsilon) |A|) \\
  & \underset{\textrm{Markov}}{\leq} & e^{-t(1+\varepsilon) |A|} \expect \left[ \prod_{s \in A} I(p \geq q) \exp( t p^{-1} X_s) \right] \\
  & \leq & e^{-t(1+\varepsilon) |A|} \expect \left[ \prod_{s \in A} I(p \geq q) h(p^{-1} X_s) \right] \\
  & \underset{\textrm{Ineq.~\ref{i:func_invariant_capped}}}{\leq} & e^{-t(1+\varepsilon) |A|} h(1)^{|A|} \\
  & \underset{\textrm{By calculus}}{\leq} & \exp\left(-\frac{n}{12} \varepsilon^2\right)
\end{eqnarray*}
where we choose $h(x) = 1+qx (e^{t/q}-1)$.
Note that $h$ is affine and it can be easily checked%
\footnote{Because the exponential function is convex and $h$ is affine, we only have to check the end points: $0, q^{-1}$.}
that it is an upper approximation of $\exp(tx)$ for $x \in [0;q^{-1}]$.
For the last step, we have to find the $t$ that produces the required bound.%
\footnote{We use $t = q \ln(1+\varepsilon)$ which is not the real optimum, but better for algebraic evaluation.}
To use these bounds, we also have to separately estimate $\prob(p < q)$.
We use a similar strategy, as in the original proof by Chakraborty et al.~\cite{chakraborty2022}, where they use $q = \frac{n}{4 |A|}$.

>>>>

An alternative is the Cram\'{e}r--Chernoff method, which is a general method to obtain tail bounds for any random variable.
It can be stated simply as $\prob(X \geq a) \leq M(t) e^{-ta}$ for all $t > 0$, where $M(t) := \expect [\exp(t X)]$ is the moment generating function of $X$.
%\footnote{The moment generating function is sometimes only defined for some values of $t$ (when the corresponding integral exists), and this is typically an interval including $0$; the inequality is only applicable wherever $M(t)$ is defined.}
%Since the probability space representing the state of the CVM algorithm is finite, $M(t)$ will always be defined in our case.
It is also possible to obtain lower tail bounds $\prob(X \leq a)$ using Cram\'{e}r--Chernoff method, which just requires estimates for $M(t)$ for $t < 0$, instead of $t > 0$.

@@

Of course, the exponential function in $M(t)$ is convex, so this is not yet the full story.
However, we can instead try to derive tail bounds for the random variable $\indicat(p \geq q) p^{-1} |\chi|$.
This leads to a similar inequality:
\begin{align}
  \expect \left[ \prod_{s \in A} I(p \geq q) h(p^{-1} X_s) \right] \leq h(1)^{|A|} \label[ineq]{i:func_invariant_capped}
\end{align}
with the new condition that $h$ needs to be non-negative and concave only on $[0;q^{-1}]$.
This then allows us to use approximate the exponential function from above with an affine function $h$ on the range $[0;q^{-1}]$, which yields tail bounds for $p^{-1} |\chi|$ under the condition $p \geq q$. As an example, the upper tail bound can be derived as follows: %TODO: should we use exp(...) everywhere instead of mixing with e^ ?
\begin{eqnarray*}
  \prob( p^{-1} |\chi| \geq (1+\varepsilon) |A| \wedge p \geq q) & \leq &
  \prob( \indicat(p \geq q) p^{-1} |\chi| \geq (1+\varepsilon) |A|) \\
  & \underset{\textrm{Markov}}{\leq} & e^{-t(1+\varepsilon) |A|} \expect \left[ \prod_{s \in A} \indicat(p \geq q) \exp( t p^{-1} X_s) \right] \\
  & \leq & e^{-t(1+\varepsilon) |A|} \expect \left[ \prod_{s \in A} \indicat(p \geq q) h(p^{-1} X_s) \right] \\
  & \underset{\textrm{Ineq.~\ref{i:func_invariant_capped}}}{\leq} & e^{-t(1+\varepsilon) |A|} h(1)^{|A|} \\
  & \underset{\textrm{By calculus}}{\leq} & \exp\left(-\frac{n}{12} \varepsilon^2\right)
\end{eqnarray*}
where we choose $h(x) = 1+qx (e^{t/q}-1)$.
Note that $h$ is affine and it can be easily checked%
\footnote{Because the exponential function is convex and $h$ is affine, we only have to check the end points: $0, q^{-1}$.}
that it is an upper approximation of $\exp(tx)$ for $x \in [0;q^{-1}]$.
For the last step, we have to find the $t$ that produces the required bound.%
\footnote{We use $t = q \ln(1+\varepsilon)$ which is not the real optimum, but better for algebraic evaluation.}
To use these bounds, we also have to separately estimate $\prob(p < q)$.
For that, we use a similar strategy, as in the original proof by Chakraborty et al.~\cite{chakraborty2022}, with $q = \frac{n}{4 |A|}$.

### 2025-02-23 paper/sec/negdep.tex
minor fixes (mostly macros or uniformity)

When we look at the subsampling step of~\cref{alg:cvm}, our invariant~\cref{i:func_invariant} imposes the following condition on the subsampling operation:
\begin{equation}\label[ineq]{i:subsample_condition}
  \int_{\mathrm{subsample}(\chi)} \prod_{s \in S} g(\indicat(s \in \tau)) \, d \tau \leq \prod_{s \in S} \expect_{\Ber(f)} [g]
\end{equation}
for all non-negative functions $g$ and any $S \subseteq \chi$---any subsampling step that satisfies this functional inequality can be used while still preserving~\cref{i:func_invariant} for the algorithm.

>>>>

When we look at the subsampling step of~\cref{alg:cvm}, our invariant~\cref{i:func_invariant} imposes the following condition on the subsampling operation:
\begin{equation}\label[ineq]{i:subsample_condition}
  \int_{\mathrm{subsample}(\chi)} \prod_{s \in S} g(\indicat(s \in \tau)) \, d \tau \leq \prod_{s \in S} \expect_{\Ber(f)} [g]
\end{equation}
for all non-negative functions $g$ and any $S \subseteq \chi$.
Any subsampling step that satisfies this functional inequality can be used while still preserving~\cref{i:func_invariant} for the algorithm.

### 2025-02-23 paper/sec/semra.tex
minor fixes (mostly macros or uniformity)

\subsection{Semantics of Randomized Algorithms}
We briefly review how reasoning about randomized algorithms works in Isabelle using the Giry monad~\cite{giry1982}.
A thorough discussion of the concept in the context of Isabelle has been written, for example, by Eberl et al.~\cite{eberl2020} and Lochbihler~\cite{lochbihler2016}.

>>>>

\subsection{Semantics of Randomized Algorithms}
We briefly review how reasoning about randomized algorithms works in Isabelle/HOL using the Giry monad~\cite{giry1982}.
A thorough discussion of the concept in the context of Isabelle/HOL has been written, for example, by Eberl et al.~\cite{eberl2020} and Lochbihler~\cite{lochbihler2016}.
